---
title: "Association of Prenatal Exposure to Air Pollutants with Select Birth Defects Using the Case-Cohort Approach"
subtitle: "Using environmental data in birth defects surveillance research"
author: "Abigail Stamm, New York State Department of Health"
date: "May 10, 2023"
output: 
  beamer_presentation:
    includes:
      in_header: preamble_beamer.tex
    latex_engine: xelatex
    keep_tex: TRUE
    slide_level: 2
bibliography: library.bib
csl: ama-citation.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr); library(kableExtra); library(rmarkdown);library(tidyverse)
```

# Introduction

## Introduction

<!-- https://airquality.wi.gov/home/map
State level: note Milwaukee & CO
County level: note Milwaukee & SO2 -->

<!-- From council:
What they are interested in is seeing an example of environmental health research that can be done with a functioning birth defects surveillance system (which Wisconsin does not now have but is hoping to develop) -->

<!-- What I emailed them

Introduction (1 min)
Birth defect data using a passive registry (3 min)
* Defining cases
* Calculating conception date
* Determining location of mother's residence
* Determining relevant maternal health & behavior (e.g. smoking) 
Environmental data (7 min)
* Considerations like location, geography, exposure, and frequency
* Air and vegetation datasets I considered and why I chose the ones I did 
* How I joined spatial data 
Example study (3 min)
* Excerpt from my dissertation looking at one of the four birth defects I studied
* Strengths and limitations of my results relative to issues noted above
Conclusion (1 min) -->

### Objectives

* Preparing birth and birth defects data
* Assessing environmental data
* Joining birth defects and environmental data
* Example analysis 

```{r functions, message = FALSE, warning = FALSE, echo = FALSE, include = FALSE}
# workaround to bypass DOH firewall
mt_subset_rev <- function (product, band, lat, lon, start = "2000-01-01", 
                           end = format(Sys.time(), "%Y-%m-%d"), km_lr = 0, km_ab = 0, 
                           site = "sitename") {
  # identify dates
  url <- paste(MODISTools:::mt_server(), product, "dates", sep = "/")
  json_dates <- httr::RETRY(verb = "GET", url = url, 
                            query = list(latitude = lat, longitude = lon), 
                            httr::write_memory(), terminate_on = c(403, 404))
  dates <- jsonlite::fromJSON(httr::content(json_dates, "text", 
                                            encoding = "UTF-8"), simplifyVector = TRUE)$dates
  
  # dates <- MODISTools::mt_dates(product = product, lat = lat, lon = lon)
  dates$calendar_date <- as.Date(dates$calendar_date)
  dates <- dates[which(dates$calendar_date <= as.Date(end) & 
                         dates$calendar_date >= as.Date(start)), ]
  breaks <- seq(1, nrow(dates), 10)
   
  # pull data
  complete_data <- lapply(band, function(band) {
    subset_data <- lapply(breaks, function(b) {
      url <- paste(MODISTools:::mt_server(), product, "subset", sep = "/")
      end_date <- dates$modis_date[b + 9]
      query <- list(latitude = lat, longitude = lon, band = band, 
                    startDate = dates$modis_date[b], endDate = end_date, 
                    kmAboveBelow = round(km_ab), kmLeftRight = round(km_lr))
      json_chunk <- httr::RETRY(verb = "GET", url = url, 
                                query = query, httr::write_memory(), 
                                terminate_on = c(403, 404))
      chunk <- jsonlite::fromJSON(httr::content(json_chunk, "text", 
                                                encoding = "UTF-8"), 
                                  simplifyVector = TRUE) 
      return(chunk)
    })
    # header set-up
    header <- subset_data[[1]][!(names(subset_data[[1]]) %in% c("header", "subset"))]
    header$site <- site
    header$product <- product
    header$start <- start
    header$end <- end
    header$cellsize <- as.character(header$cellsize)
    header$complete <- !any(unlist(lapply(subset_data, is.null)))
    
    # data
    subset_data <- do.call("rbind", lapply(subset_data, function(x) x$subset))
    pixels <- do.call("rbind", subset_data$data)
    colnames(pixels) <- seq_len(ncol(pixels))
    subset_data <- cbind(subset_data[, !(names(subset_data) %in% 
                                           "data")], pixels)
    subset_data <- stats::reshape(subset_data, 
                                  varying = grep("[0-9]",  names(subset_data)), 
                                  direction = "long", timevar = "pixel", 
                                  v.names = "value")
    subset_data <- subset_data[, !(names(subset_data) %in% "id")]
    subset_data <- data.frame(header, subset_data, stringsAsFactors = FALSE)
    subset_data <- subset_data[, !(names(subset_data) %in% "band.1")]
  
  })
  return(do.call("rbind", complete_data))
}


```


```{r dataprep, message = FALSE, warning = FALSE, echo = FALSE, include = FALSE}
cny <- tigris::counties(state = "NY")
tny <- tigris::tracts(state = "NY", year = 2010)

border_counties <- c("Erie", "Oswego", "Chautauqua", "Jefferson", "Monroe", 
                     "Wayne", "Cayuga", "Suffolk", "Nassau", "New York", 
                     "Queens", "Kings", "Richmond", "Bronx", "Niagara",
                     "Orleans", "St. Lawrence")
l <- list()

for (i in border_counties) {
  l[[i]] <- tigris::area_water(state = "NY", county = i) |>
    dplyr::filter(FULLNAME %in% c("Lk Ontario", "Atlantic Ocean", "Lk Erie",
                                  "East Riv", "Hudson Riv", "Harlem Riv") | 
                    grepl("Bay$|Sound$", FULLNAME) | MTFCC == "H2053" |
                    (is.na(FULLNAME) & MTFCC == "H2030"))
}

wny <- dplyr::bind_rows(l) |> dplyr::filter(AWATER > 10000000)

a <- readr::read_csv("data/airmonitorlocations.csv") |>
  sf::st_as_sf(coords = c("Longitude", "Latitude"), crs = sf::st_crs(t))

url <- paste(MODISTools:::mt_server(), "MOD13Q1", "bands", sep = "/")
bands <- jsonlite::fromJSON(readr::read_file(url))$band |>
  dplyr::filter(grepl("NDVI", band))

cwi <- tigris::counties(state = "WI") |> 
    dplyr::filter(NAME == "Dane")

twi <- tigris::tracts(state = "WI", year = 2010)
modis <- mt_subset_rev(product = "MOD13Q1", lat = 43.1, lon =  -89.4,
                       band = bands$band, site = "madison", 
                       km_lr = 20, km_ab = 20, # km buffers: east-west, north-south
                       start = "2023-04-01", end = "2023-04-30")

wmap <- MODISTools::mt_to_terra(df = modis, reproject = TRUE)

nlcd <- FedData::get_nlcd(template = cwi, label = "wi", year = 2011, 
                          dataset = "landcover", landmass = "L48")
# ncol <- FedData::nlcd_colors()
npal <- FedData::pal_nlcd()
nid <- c(11, 21:24, 41:43, 71, 81:95)
ntemp <- npal[npal$ID %in% nid, ]


nlct <- raster::projectRaster(nlcd, crs = sp::CRS("EPSG:4269"), 
                              method = "ngb")

twi2 <- sf::st_crop(twi, c(xmin= -89.5, ymin = 43, xmax = -89.3, ymax = 43.15)) |>
  dplyr::mutate(pct_water = AWATER10 / (ALAND10 + AWATER10)) |> 
  dplyr::filter(pct_water < .5)
twi3 <- sf::st_crop(twi, c(xmin= -89.5, ymin = 43, xmax = -89.3, ymax = 43.15)) |>
  dplyr::mutate(pct_water = AWATER10 / (ALAND10 + AWATER10)) |> 
  dplyr::filter(pct_water > .5)

```

# Birth defects data

## Defining cases

### Relevant variables (from birth defects registry)

* British Pediatric Association (BPA) code
* Birth defects description (narrative)

### Concerns

* May only have access to BPA or ICD code
* Code and narrative may contradict
* Include or exclude cases with multiple birth defects?

<!-- ... as well as year, residency, whatever else is relevant; 
I assigned case by code, but excluded hereditary conditions based on code and narrative -->

## Calculating conception date

### Relevant variables (from vital records)

* Last menstrual period (LMP)
* Clinical gestational age (with birthdate)

### Concerns

* Data may be missing
* Dates may not agree
* Dates may be impossible

<!-- I assigned conception date by reasonable LMP first (<350 days) then by birthdate minus clinical age (if <350 days) and excluded the rest; 
no conception date = couldn't join to air pollution data -->

## Determining location of mother's residence

### Relevant variables (from vital records)

* Mother's residence address at birth (street, city, ZIP code, state)

### Concerns

* Address may not geocode
* May have access to only part of address
* State, city, ZIP code may disagree

<!-- I geocoded what I could (prioritized ZIP code over city and state), then used ZIP code centroid for the few that did not geocode, and those without ZIP code I dropped = couldn't join to environmental data -->

## Determining maternal health & behavior

### Relevant variables (from vital records)

* Mother's smoking & alcohol use
* Mother's prenatal care
* Mother's gestational diabetes & body-mass index

### Concerns

* Data may be missing
* Data may be impossible or unreliable
* No data on father's health & behavior

<!-- I included only smoking and my model dropped missings -->

# Environmental data

## Environmental considerations

*Geography:* How wide is coverage?

*Resolution:* What is the resolution level?

*Exposure:* What dosage assumptions will I make?

*Frequency:* Are data daily, monthly, annual?

*Availability:* Are data accessible?

## Air pollution: monitoring stations

\begincols{}
\column{0.5\textwidth}

\small

* *Geography:* Mostly urban
* *Resolution:* Points far apart
* *Exposure:* Measured values
* *Frequency:* Daily or every few days
* *Availability:* https://www.airnow.gov/

\normalsize

\column{0.5\textwidth}

```{r airmonitor, message = FALSE, warning = FALSE}
graphics::par(mar=c(.1,.1,.1,.1))

plot(cny$geometry, border = "darkgrey", col = "ivory")
plot(wny$geometry, add = TRUE, border = "mediumblue", col = "lightblue")
plot(a$geometry, add = TRUE, pch = 20, col = "red")
```

\centering
\footnotesize
Map of New York air pollution monitors (Source: New York State Department of Environmental Conservation)
\normalsize
\stopcols

## Air pollution: Downscaler model

\begincols{}
\column{0.5\textwidth}

\small

* *Geography:* Nationwide
* *Resolution:* Census tract 
* *Exposure:* Estimated values
* *Frequency:* Daily
* Uses topography, weather, pollution sources
* *Availability:* https://www.epa.gov/hesc/rsig-related-downloadable-data-files

\normalsize

\column{0.5\textwidth}

```{r airmodel, message = FALSE, warning = FALSE}
graphics::par(mar=c(.1,.1,.1,.1), mgp = c(0, 0, 0), xpd = TRUE)

plot(tny$geometry, border = "darkgrey", col = "ivory")
plot(wny$geometry, add = TRUE, border = "mediumblue", col = "lightblue")

```

\centering
\footnotesize
Map of New York Census tracts
\normalsize
\stopcols

## Green space: Normalized Difference Vegetation Index (NDVI)

\begincols{}
\column{0.5\textwidth}

\small

* *Geography:* Global
* *Resolution:* 250 m$^2$ grid
* *Exposure:* Measured values
* *Frequency:* Every 16 days
* Cloud cover, unclear land types
* *Availability:* R package MODISTools

\normalsize

\column{0.5\textwidth}

```{r ndvi, fig.width = 3, fig.height = 2.5, message = FALSE, warning = FALSE}
graphics::par(mar=c(.1,.1,.1,.1), mgp = c(0, 0, 0), xpd = TRUE)
# mar 

my_window <- raster::extent(-89.5, -89.25, 43.0, 43.15)
raster::plot(my_window, col=NA, axes = FALSE, # legend = FALSE, 
             xlab = "", ylab = "")

raster::plot(wmap, xlim = c(-89.5, -89.3), ylim = c(43.0, 43.15), axes = FALSE,
         legend.args=list(side = 4, line = 0, cex = 2), 
         # legend.width = 1, legend.shrink = 1,
         add = TRUE) # 

plot(twi2$geometry, border = "black", lwd = 1, add = TRUE, legend = FALSE) # 
```

\centering
\footnotesize
NDVI grid and Census tracts around Madison, WI
\normalsize
\stopcols

## Green space: National Land Cover Database (NLCD)

\begincols{}
\column{0.5\textwidth}

\small

* *Geography:* Nationwide
* *Resolution:* 30 m$^2$ grid
* *Exposure:* Categorized land types
* *Frequency:* 3-year estimates
* *Availability:* R package FedData

\normalsize

\column{0.5\textwidth}

```{r nlcd, message = FALSE, warning = FALSE}
graphics::par(mar=c(.1,.1,.1,.1), mgp = c(0, 0, 0), xpd = TRUE)

my_window <- raster::extent(-89.6, -89.3, 43.0, 43.15)
raster::plot(my_window, col=NA, axes = FALSE, # legend = FALSE, 
             xlab = "", ylab = "")

raster::plot(nlct, xlim = c(-89.5, -89.3), ylim = c(43.0, 43.15), 
         col = npal$Color, axes = FALSE, legend = FALSE,
         # legend.args=list(side = 1, line = 0), 
         xaxt = 'n', yaxt = 'n', horizontal = TRUE, add = TRUE)


plot(twi2$geometry, border = "black", lwd = 1, add = TRUE) # 
legend(x='bottomleft', legend = ntemp$Class, fill = ntemp$Color, cex = 1.5, 
       bg = "white")

```

\centering
\footnotesize
NLCD grid and Census tracts around Madison, WI
\normalsize
\stopcols

## Environmental datasets selected

### Project requirements

* Statewide
* Daily or weekly PM$_{2.5}$ and ozone measures
* Differentiate grasses, trees, and water

### Final selections

* Air pollution: Downscaler modeled data, 2002-2015
* Green space: National Land Cover Database, 2011

# Joining spatial data

## Join birth defect and air pollution data

\begincols{}
\column{0.5\textwidth}

1. Plot mothers' residences and join to Census tracts
2. Join mothers' residences to air pollution data on both Census tract and conception date

\column{0.5\textwidth}

```{r pointtract, message = FALSE, warning = FALSE, fig.width=2.5, fig.height=1.5}
pts <- data.frame(name = c("Capitol Building", "Waisman Center"), 
                  lon = c(-89.384, -89.433), 
                  lat = c(43.075, 43.079),
                  geoid10 = c("55025001704", "55025003200"), 
                  cdate = c("2019-04-01", "2019-04-23"))

graphics::par(mar=c(.1,.1,.1,.1), mgp = c(0, 0, 0), xpd = TRUE)

my_window <- raster::extent(-89.45, -89.35, 43.05, 43.1)
raster::plot(my_window, col=NA, axes = FALSE, # legend = FALSE, 
             xlab = "", ylab = "")
plot(twi3$geometry, border = "transparent", col = "cornflowerblue", 
     add = TRUE) # 
plot(twi2$geometry, border = "black", lwd = 1, add = TRUE) # 
points(pts$lon, pts$lat, pch = 20, col = "red")
```

\centering
\footnotesize
Waisman Center and Capitol Building, Madison, WI
\normalsize

\stopcols

$~$

```{r birthtable}
births <- pts |> dplyr::select(name, geoid10, cdate)

births |> kable(booktabs = TRUE, format = "latex", 
                caption = NA, 
                col.names = c("Name", "Census tract", "Conception date")) 
```

## Join birth defect and green space data

\begincols{}
\column{0.5\textwidth}

1. Calculate buffers around mothers' residences and join to green space
2. Calculate proportion of buffers with each green space type

\column{0.5\textwidth}

```{r buffergrid, message = FALSE, warning = FALSE, fig.width=3, fig.height=3}
graphics::par(mar=c(.1,.1,.1,.1), mgp = c(0, 0, 0), xpd = TRUE)

my_window <- raster::extent(-89.45, -89.35, 43.05, 43.1)
raster::plot(my_window, col=NA, axes = FALSE, # legend = FALSE, 
             xlab = "", ylab = "")

# plot(twi2$geometry, border = "black", lwd = 1, add = TRUE) # 
raster::plot(nlct, xlim = c(-89.45, -89.35), ylim = c(43.05, 43.1), 
         col = npal$Color, axes = FALSE, legend = FALSE,
         # legend.args=list(side = 1, line = 0), 
         xaxt = 'n', yaxt = 'n', horizontal = TRUE, add = TRUE)

points(pts$lon, pts$lat, pch = 20, col = "red")

plotrix::draw.circle(x=pts$lon[1], y=pts$lat[1], radius=.005)
plotrix::draw.circle(x=pts$lon[2], y=pts$lat[2], radius=.005)

nid <- c(11:42, 81, 90:95)
nt2 <- ntemp[ntemp$ID %in% nid, ]
legend(x='bottomleft', legend = nt2$Class, fill = nt2$Color, cex = .5, 
       bg = "white")

```

\centering
\footnotesize
NLCD grid with Waisman Center and Capitol Building, Madison, WI
\normalsize

\stopcols

# My study

## Analysis

### Purpose

To determine the most vulnerable time between one month before pregnancy and the end of the third month of pregnancy.

### Design

Case-cohort study, NYS excluding New York City, 2002-2015

### Model

Regression with weekly exposures

## Example results

```{r resultsetup, include = FALSE}
devtools::load_all("C:/Users/ajstamm/Documents/GitHub/Coding/apcmpkg/R")

draw_forest <- function(data, legend = FALSE, miny = 0.5, maxy = 1.5) {
  my_x <- max(data$lag)
  data$lag <- ifelse(grepl("o3", data$air),
                     data$lag - 0.08, data$lag + 0.08)

  #mean profiles in the same graphs
  mp <- ggplot2::ggplot(data, ggplot2::aes(x=lag, y=rr, group = air,
                                           fill = air, col = air)) +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "bottom",
          text = ggplot2::element_text(size = 14),
          plot.title = ggplot2::element_text(size = 14),
          axis.text = ggplot2::element_text(size = 14),
          legend.text = ggplot2::element_text(size = 14)) +
    ggplot2::geom_hline(yintercept = 1) +
    ggplot2::geom_vline(xintercept = seq(from = -4, to = my_x, by = 4)) +
    ggplot2::geom_line() +
    ggplot2::geom_point(size = 2.5) +
    ggplot2::labs(x = "Week of pregnancy", y = "Risk ratio") + #
    ggplot2::guides(shape = "none") +
    ggplot2::scale_color_manual(labels = c("Ozone", "PM2.5"),
                       values = c("blue", "red"),
                       name = "Air pollutant") +
    ggplot2::scale_fill_manual(labels = c("Ozone", "PM2.5"),
                      values = c("blue", "red"),
                      name = "Air pollutant") +
    ggplot2::geom_errorbar(ggplot2::aes(ymin = cilow, ymax = cihigh),
                           width = 0.1) +
    ggplot2::scale_x_continuous(breaks = seq(from = -4, to = my_x, by = 2)) +
    ggplot2::scale_y_continuous(breaks = seq(from = miny, to = maxy,
                                             by = (maxy - miny) / 4),
                                limits = c(miny, maxy))
  return(mp)
}

draw_forest_rev <- function(data, defect_abbr, defect_name = "", green = "") {
  d <- dplyr::filter(data, defect == defect_abbr, !lag == "cum")
  d <- dplyr::mutate(d, lag = (as.numeric(gsub("lag", "", lag)) * -1) +
                       (lags - 4))
  min <- floor(min(d$cilow)*10)/10
  max <- ceiling(max(d$cihigh)*10)/10
  # read each plot
  dm16 <- dplyr::filter(d, grepl("mean", air), lags == 16)
  m16 <- draw_forest(data = dm16, miny = min, maxy = max, 
                     legend = TRUE)

  main <- paste(defect_name, "16-week model: \n", 
                "Risk ratios by week of pregnancy, ",
                "300m buffer with grasses and trees")
  
  footer <- paste("Note: Models were adjusted for maternal education level,",
                  "maternal smoking, tract-level median income,",
                  "conception season, and the indicated green space variable. ", 
                  "Week 0 was the week of conception. ",
                  "Week 12 was the end of the first trimester of pregnancy. ",
                  "Risk ratio applies to change in risk from the previous week",
                  "given a 10-unit increase in the air pollutant",
                  "(ppb for ozone and µg/m\U00B3 for PM2.5)",
                  "over two standard deviations above the mean.")

  foot <- gridtext::textbox_grob(footer, width = unit(9, "in"), r = unit(0, "pt"),
                                 # x = unit(0.5, "npc"), y = unit(0.7, "npc"),
                                 gp = grid::gpar(justify = "left", fontsize = 14),
                                 padding = unit(c(10, 20, 10, 10), "pt"),
                                 margin = unit(c(10, 20, 10, 10), "pt"),
                                 valign = 0.5)

  return(gridExtra::grid.arrange(ggplot2::ggplotGrob(m16),
                                 foot, # widths = 6,
                                 layout_matrix = cbind(1:2),
                                 ncol = 1, # top = main, bottom =
                                 heights = c(1.9, 0.7)))

  # return grid directly to avoid tablegrob comment
  # return(gridExtra::grid.arrange(x, y, ncol = 1))
}

```

Craniosynostosis models: Risk ratios by week of pregnancy with grasses and trees

```{r cp3gtbar, fig.width=9, fig.height=5, warning = FALSE, message = FALSE}
d <- readr::read_csv("data/model2gt_rr_all.csv")


draw_forest_rev(data = d, defect_abbr = "cs", green = "gt", 
                defect_name = "Craniosynostosis")

```

## Summary

### Greatest increases in risk

```{r resultstable}
bds <- data.frame(bd = c("Clubfoot", 
                         "Cleft lip w/wo cleft palate", 
                         "Cleft palate", 
                         "Craniosynostosis"), 
                  o3 = c("Pre-conception", 
                         "Pre-conception", 
                         "Pre-conception", 
                         "Month 1"), 
                  pm = c("Month 1-2", 
                         "Month 1", 
                         "Month 2", 
                         "Pre-conception"))

bds |> kable(booktabs = TRUE, format = "latex", 
             col.names = c("Birth defect", "O3 exposure", "PM2.5 exposure"), 
             caption = NA)  # |>

```

### Observation

Results varied depending on air pollutant measure and model settings

# Contact

## Contact Information

### Abigail Stamm

Email: abigail.stamm@health.ny.gov   
Dissertation: https://github.com/ajstamm/apcmpkg

### NYS Birth Defects Registry

Email: BDR@health.ny.gov   
Contact: Michele Herdt or Amanda St Louis   
Website: https://health.ny.gov/diseases/congenital_malformations/   

